\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-01-17 papers/methods/paper.tex
Paper updates

Traditional $\chi^2$ analyses such as that found in \citet{Kowalski2008, Conley2011, Betoule2014}, minimise the difference in distance modulus between the cosmologically predicted values $\mu_C$ and the observed distance modulus $\mu_{\rm obs}$, shown respectively below:
\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right]\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}} \\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B
\end{align}
The minimising function is then given as
\begin{equation}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C)
\end{equation}
where $C^{-1}$ is an uncertainty matrix that combined the uncertainty from the SALT2 fits, intrinsic dispersion, calibration, dust, peculiar velocity and many other factors (see \citet{Betoule2014} for a review). The benefit this analysis methodology provides is speed - for samples of hundreds of supernovae, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with two costs. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of uncertainty Gaussianity. Secondly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

@@

Like most of the BHM methods introduced previously, we work from the summary statistics as well, where each observed supernova has an apparent magnitude $\hat{m_B}$, stretch $\hat{x_1}$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$. Given machine learning classifiers to type supernovae, we will also have a probability of being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m_B}, \hat{x_1}, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

@@

We model the underlying population and intrinsic dispersion together as a redshift-dependent multivariate skew normal for each survey. Following {\rubin} we allow the mean colour and stretch to vary over redshift, anchoring four equally spaced redshift nodes spanning the redshift range of each survey, linearly interpolating between the nodes. These nodes are represented as $\langle x_1^i \rangle$ and $\langle c^i \rangle$. Both the colour and stretch means are modelled with normal priors. The correlation between the magnitude, stretch and colour populations is represented in a correlation matrix $\rho$, which is treated with an LKJ prior. Skewness is fixed to 0 for the absolute magnitude and stretch, but left free for colour. We parameterise the standard skew normal skewness $\alpha_c$ by sampling $\delta_c = \alpha_c / \sqrt{1 + \alpha_c^2}$ which itself is given a uniform prior $\mathcal{U}(0,0.98)$.

The width of the population, represented by the vector $\lbrace \sigma_{M_B}, \sigma_{x_1}, \sigma_c \rbrace$ is subject to Cauchy priors, however are sampled in log space for efficiency in sampling close to zero.

As such, the only constant between survey populations is the absolute magnitude $M_B$, with the skewness, redshift-dependent means, width and correlations fit individually for each survey.

@@

We tested two models against simulated supernova sets; $\beta(c) = \beta_0 + \beta_1 c$ and $\beta(z) = \beta_0 + \beta_1 z$. See Section \ref{sec:simdes} for details on simulation generation. We found for both models that non-zero values for $\beta_1$ are preferred (even with constant $\beta$ used in simulation) due to severe degeneracy with selection effects. This degeneracy resulted in a significant bias in recovered cosmology, and so in our final model we continue to adopt the constant $\alpha$ and $\beta$ found in traditional analyses.

@@

It is now well known that host galaxy environment has a significant effect on supernova properties. The latest sample of over 1300 spectroscopically confirmed Type Ia supernovae show $>5\sigma$ evidence for correlation between host mass and luminosity \citep{Uddin2017}. The traditional correction, as employed in analyses such as \citet{Suzuki2012} and \citet{Betoule2014} invoke a step function such that $\Delta M = 0.08 \mathcal{H}(\log(M) - 10))$, where $\mathcal{H}$ is the Heaviside step function and $M$ is the galaxy mass in solar masses. The scale of this step function varies from analysis to analysis, with the 0.08 value shown previously sourced from \cite{Sullivan2010} and used in \citet{Betoule2014}. In this work we adopt the model used in {\rubin}, which follows the work from \citet{Rigault2013}, such that we introduce two parameters to incorporate a redshift-dependent host galaxy mass correction:
\begin{equation}
\Delta M = \delta(0) \left[ \frac{1.9\left(1 - \frac{\delta(0)}{\delta(\infty)}\right)  }{0.9 + 10^{0.95z}} + \frac{\delta(0)}{\delta(\infty)}\right]
\end{equation}
We also take flat priors on the parametrisation $\delta(0)$, $\delta(0)/\delta(\infty)$.

@@

This method of adjusting the observed summary statistics is used throughout the traditional and BHM analyses, however it is normally constrained to band systematics. That is, each band for each survey has two systematics associated with it - the calibration uncertainty and the filter wavelength uncertainty. We include these in our approach, in addition to including HST Calspec calibration uncertainty, 10 SALT2 model systematic uncertainties, a dust systematic, a global redshift bias systematic and also the systematic peculiar velocity uncertainty. This gives thirteen global systematics shared by all surveys, plus two systematics per band in each survey. Denoting $\eta \equiv \lbrace m_B, x_1, c \rbrace$, our initial conditional likelihood for our observed summary statistics shown in Equation \eqref{eq:pop} becomes
\begin{align}
P(\hat{\eta}, \frac{\partial \hat{\eta}}{\partial \Z_i} | \eta, \delta \Z_i, C) = \mathcal{N}(\hat{\eta} + \delta \Z_i \frac{\partial \hat{\eta}}{\partial \Z_i}|\eta,\cov). \label{eq:pop2}
\end{align}

\subsubsection{Selection Effects}
\label{sec:selection}
Our treatment of selection effects is to incorporate selection efficiency into our model, rather than relying on simulation-driven data corrections. As such, we need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as
\begin{align}
\mathcal{L} &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As the model so far described in previous sections describes $P(D|\theta)$, and we wish to formulate a function $P(S|{\rm data},\theta)$ that describes the chance of an event being successfully observed, we rearrange the likelihood in terms of those functions and find
\begin{align}
\mathcal{L} &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}
where the denominator represents an integral over all potential data. Full derivation of this can be found in Appendix \ref{app:selection}. As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by HMC. We therefore simplify the integral and approximate the selection effects in apparent magnitude and redshift space independently, such that the denominator, denoted now $w$ for simplicity, is given as
\begin{align}
w &= \int  \left[ \int P(S|m_B) P(m_B | z, \theta)\, d m_B \right] P(S|z) P(z|\theta)\, dz. \label{eq:w1}
\end{align}
We apply two further approximations similar to those made in {\rubin} -- that the redshift distribution of the observed supernova reasonably well sample the $P(S|z)P(z|\theta)$ distribution, and that the survey colour and stretch populations can be treated as Gaussian for the purposes of this integral. The population $P(m_B | z, \theta)$ thus becomes $\mathcal{N}(m_B|m_B^*(z), \sigma^*_{m_B})$, where
\begin{align}
m_B^*(z) &= \langle M_B \rangle + \mu(z) - \alpha \langle x_1(z) \rangle + \beta \langle c(z) \rangle \\
\sigma^*_{m_B} &= \sigma_{M_B}^2 + (\alpha \sigma_{x_1})^2 + (\beta \sigma_c)^2 \notag\\ 
&\quad\quad + 2(\beta \sigma_{M_B,c} - \alpha \sigma{m_B,x_1} - \alpha \beta \sigma_{x_1,c})
\end{align}
What then remains is determining the functional form of $P(S|m_B)$. For the treatment of the DES, we find that the error function which smoothly transitions from some constant efficiency down to zero is sufficient. This similar result has been found by many past surveys \citep{Dilday2008, Barbary2010, Perrett2012, Graur2013, Rodney2014} {\red THESE CITATIONS EXACTLY MIRROR RUBIN2015, NOT SURE IF I SHOULD CHANGE THEM}. For our treatment of the low redshift surveys included {\red Introduce which lowz surveys earlier on} we simplify the model fitting and group all low redshift surveys together. This requires a more complicated selection function than the error function used for DES, however is reasonably well fit with a skew normal.

The selection functions were fit to apparent magnitude efficiency ratios calculated from SNANA simulations by calculating an efficiency ratio as a function of apparent magnitude. Uncertainty of the Malmquist bias (entering both through statistical uncertainty from finite sized simulations in the efficiency ratio and the discrepancy between the analytic approximation and non-analytic simulation results) is incorporated into the fitting for the analytic approximation. Uncertainty is uniformly added to the efficiency ratio until the reduced $\chi^2$ of the analytic fit reached $1$, allowing us to extract an uncertainty covariance matrix for our analytic fits.

The DES survey is thus characterised by mean fit values $\mu_{\rm DES} = 22.3$, $\sigma_{\rm DES} = 0.7$, with the combined low redshift surveys characterised by $\mu_{\rm LowZ} = 14.1$, $\sigma_{\rm LowZ} = 1.36$, $\alpha_{\rm LowZ} = 3.8$.

With the well sampled approximation as specified previously, we can remove the redshift integral in Eq \eqref{eq:w1} and replace it with a correction for each observed supernova. For the error function (DES) and skew normal selection (LowZ) functions respectively, this correction becomes
\begin{align}
w_{\rm DES} &= \Phi^c\left(  \frac{m^*_B - \mu_{\rm DES}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm DES}^2}}  \right) \label{eq:seldes}\\
w_{\rm LowZ} &= 2\mathcal{N}\left( \frac{m_B^* - \mu_{\rm LowZ}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm LowZ}^2}}\right) \times \notag\\
&\quad\quad \Phi\left(  \frac{{\rm sign}(\alpha_{\rm LowZ})(m_B^* - \mu_{\rm LowZ})}{\frac{\sigma_{m_B}^{*2} + \sigma^2_{\rm LowZ}}{\sigma^2_{\rm LowZ}} \sqrt{\frac{\sigma^2_{\rm LowZ}}{\alpha^2_{\rm LowZ}} + \frac{\sigma_{m_B}^{*2} \sigma^2_{\rm LowZ}}{\sigma_{m_B}^{*2} + \sigma^2_{\rm LowZ}}} }\right), \label{eq:sellowz}
\end{align}
where $\Phi$ represents the normal CDF function, and $\Phi^c$ the complimentary CDF.

@@

In order to verify our model we run it through several tests. First, we validate on toy models, verifying that there is not significant cosmological bias in data sets more constraining than the DES spectroscopic sample. We then also generate

@@

We generate simple toy data to validate the basic premise of the model. For both DES and LowZ data we draw from an underlying $M_B$, $x_1$, $c$ population and translate into apparent magnitude space using $\Omega_m = 0.3$, $\alpha= 0.14$ and $\beta = 3.1$. Masses are randomly drawn from the interval 0 to 1, and a mass correction with $\delta(0) = 0.08$ and $\delta(0)/\delta(\infty) =0.5$ included. Independent observational errors of $0.04$, $0.2$, $0.03$ on $m_B$, $x_1$ and $c$ (following the mean uncertainty for DES SNANA simulations) are added to create the observables. The results are then passed through the selection effects, where each supernova is only selected based on $P(S|m_B)$, using a skew normal function for the LowZ supernovae and error function for the DES-like supernovae. We draw from each survey simulation until we have 300 LowZ supernovae and 700 DES-like supernovae, representing a statistical sample of greater power than the estimated 250 supernovae for the DES spectroscopic analysis.

The above represents one realisation of data. We test three models (Flat $\Lambda$CDM, Flat $w$CDM, $\Lambda$CDM) with 300 realisations and find recovery of underlying cosmology without significant bias. Combined posterior surfaces of 22 realisations for Flat $wCDM$ fits are shown in Figure \ref{fig:simple_w}.

%\begin{figure}
%	\begin{center}
%		\includegraphics[width=\columnwidth]{approximate_simple_test_cosmo.pdf}
%	\end{center}
%	\caption{Posterior surfaces for 22 realisations of supernova data with the Flat $\Lambda$CDM model.}
%	\label{fig:simple_om}
%\end{figure}
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{approximate_simple_test_w_cosmo.pdf}
	\end{center}
	\caption{Posterior surfaces for 22 realisations of supernova data with the Flat $w$CDM model.}
	\label{fig:simple_w}
\end{figure}
%\begin{figure}
%	\begin{center}
%		\includegraphics[width=\columnwidth]{approximate_simple_test_ol_cosmo.pdf}
%	\end{center}
%	\caption{Posterior surfaces for 22 realisations of supernova data with the $\Lambda$CDM model.}
%	\label{fig:simple_ol}
%\end{figure}

@@

Early analyses often treated intrinsic dispersion simply as scatter in the underlying absolute magnitude of the underlying population, but recent analyses require more a more sophisticated approach. In our development of this model and tests of intrinsic dispersion, we analyse the effects of two different scatter models. The first model is the \citet[][hereafter denoted the {\gten} scatter model]{Guy2010}, which models intrinsic scatter with a 70\% contribution from coherent variation and 30\% from chromatic variation. The second model, denoted the {\celeven} model is sourced from \citet{Chotard2011} and has variation with 25\% contribution from coherent scatter and 75\% from chromatic variation. We also test a `Combined' smear model, which is simply to draw equal samples from the {\celeven} and {\gten} models. It is against this combined sample that we wish to rigorously validate our model.

@@

\begin{table}
	\centering
	\caption{Tested population distributions, where the {\sk} LowZ distribution is formed as sum of two bifurcated Gaussians, with the mean and spread of each component given respectively.}
	\label{tab:dist}
	\begin{tabular}{l|llll}
		\hline
		Model & $\langle x_1 \rangle$ & $\sigma_{x_1}$ &  $\langle c \rangle$ & $\sigma_c$  \\
		\hline
		Gaussian LowZ &  $0.0$ & $1.0$ & $0.0$ & $0.1$ \\
		Gaussian DES  &  $0.0$ & $1.0$ & $0.0$ & $0.1$  \\
		{\sk} LowZ    &  $0.55$ \& $-1.5$ & $^{+0.45}_{-1.0}$ \& $^{+0.5}_{-0.5}$ & $-0.055$ & $^{+0.15}_{-0.023}$ \\
		{\sk} DES     & $0.973$ & $^{+0.222}_{1.472}$ & $-0.054$  &  $^{+0.101}_{0.043}$ \\
		\hline
	\end{tabular}
\end{table}

@@

To forecast the results for the 3 year DES spectroscopic supernovae sample, we generate 100 realisations of datasets, each comprising 300 low redshift supernovae and 250 DES-like supernovae. We fit these through the three cosmological models - Flat $\Lambda$CDM, $\Lambda$CDM and $w$CDM, however focus mostly on the $w$CDM model as it is the primary scientific concern of the Dark Energy Survey. We run all three cosmological models with systematics enabled and disabled to explicitly show the uncertainty included from systematic contribution.

>>>>

Traditional $\chi^2$ analyses such as that found in \citet{Kowalski2008, Conley2011, Betoule2014}, minimise the difference in distance modulus between the cosmologically predicted values $\mu_C$ and the observed distance modulus $\mu_{\rm obs}$, shown respectively below:
\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right]\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}} \label{eq:mu0}\\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B \label{eq:mu1}
\end{align}
The minimising function is then given as
\begin{equation}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C)
\end{equation}
where $C^{-1}$ is an uncertainty matrix that combined the uncertainty from the SALT2 fits, intrinsic dispersion, calibration, dust, peculiar velocity and many other factors (see \citet{Betoule2014} for a review). The benefit this analysis methodology provides is speed - for samples of hundreds of supernovae, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with two costs. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of uncertainty Gaussianity. Secondly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

@@

Like most of the BHM methods introduced previously, we work from the summary statistics as well, where each observed supernova has an apparent magnitude $\hat{m_B}$, stretch $\hat{x_1}$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^10$ solar masses. Given machine learning classifiers to type supernovae, we will also have a probability of being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m_B}, \hat{x_1}, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

@@

We model the underlying population and intrinsic dispersion together as a redshift-dependent multivariate skew normal for each survey. Following {\rubin} we allow the mean colour and stretch to vary over redshift, anchoring four equally spaced redshift nodes spanning the redshift range of each survey, linearly interpolating between the nodes. These nodes are represented as $\langle x_1^i \rangle$ and $\langle c^i \rangle$. Both the colour and stretch means are modelled with normal priors. The correlation between the magnitude, stretch and colour populations is represented in a correlation matrix $\rho$, which is treated with an LKJ prior. Skewness is fixed to 0 for the absolute magnitude and stretch, but left free for colour. We parameterise the standard skew normal skewness $\alpha_c$ by sampling $\delta_c = \alpha_c / \sqrt{1 + \alpha_c^2}$ which itself is given a uniform prior $\mathcal{U}(0,0.98)$. The width of the population, represented by the vector $\lbrace \sigma_{M_B}, \sigma_{x_1}, \sigma_c \rbrace$ is subject to Cauchy priors, however are sampled in log space for efficiency in sampling close to zero.

As such, the only constant between survey populations is the absolute magnitude $M_B$, with the skewness, redshift-dependent means, width and correlations fit individually for each survey. The probability for a supernova to have true values $M_B$, $x_1$, $c$ given the underlying population is thus given as
\begin{align}
P(M_B, x_1, c, z | \langle M_B \rangle, \langle x_1(z) \rangle, \langle c(z) \rangle, \rho, \sigma_{m_B}, \sigma_{x_1}, \sigma_c, \alpha_c) = \notag \\
\mathcal{N}^{\rm{skew}}(\lbrace M_B, x_1, c \rbrace | \lbrace \langle M_B \rangle, \langle x_1(z) \rangle, \langle c(z) \rangle \rbrace, \rho \cdot (\vec{\sigma} \vec{\sigma}^T), 
\vec{\alpha}) \label{eq:l1},
\end{align}
where $\vec{\alpha} = \lbrace 0, 0, \alpha_c \rbrace$ and $\vec{\sigma} = \lbrace \sigma_{m_B}, \sigma_{x_1}, \sigma_c \rbrace$.

@@

We tested two models against simulated supernova sets; $\beta(c) = \beta_0 + \beta_1 c$ and $\beta(z) = \beta_0 + \beta_1 z$. See Section \ref{sec:simdes} for details on simulation generation. We found for both models that non-zero values for $\beta_1$ are preferred (even with constant $\beta$ used in simulation) due to severe degeneracy with selection effects. This degeneracy resulted in a significant bias in recovered cosmology, and so in our final model we continue to adopt the constant $\alpha$ and $\beta$ found in traditional analyses. As such, our calculation of distance modulus $\mu$ mirrors that found in Equations \eqref{eq:mu0} and \eqref{eq:mu1}.

@@

It is now well known that host galaxy environment has a significant effect on supernova properties. The latest sample of over 1300 spectroscopically confirmed Type Ia supernovae show $>5\sigma$ evidence for correlation between host mass and luminosity \citep{Uddin2017}. The traditional correction, as employed in analyses such as \citet{Suzuki2012} and \citet{Betoule2014} invoke a step function such that $\Delta M = 0.08 \mathcal{H}(\log(M) - 10))$, where $\mathcal{H}$ is the Heaviside step function and $M$ is the galaxy mass in solar masses. The scale of this step function varies from analysis to analysis, with the 0.08 value shown previously sourced from \cite{Sullivan2010} and used in \citet{Betoule2014}. In this work we adopt the model used in {\rubin}, which follows the work from \citet{Rigault2013}, such that we introduce two parameters to incorporate a redshift-dependent host galaxy mass correction:
\begin{equation}
\Delta M = \delta(0) \left[ \frac{1.9\left(1 - \frac{\delta(0)}{\delta(\infty)}\right)  }{0.9 + 10^{0.95z}} + \frac{\delta(0)}{\delta(\infty)}\right]
\end{equation}
We also take flat priors on the parametrisation $\delta(0)$, $\delta(0)/\delta(\infty)$. With this correction, our calculation of absolute magnitude becomes
\begin{align}
M_B = m_B - \mu(z) - \alpha x_1 + \beta c - \Delta M \cdot m. \label{eq:l2}
\end{align}

@@

This method of adjusting the observed summary statistics is used throughout the traditional and BHM analyses, however it is normally constrained to band systematics. That is, each band for each survey has two systematics associated with it - the calibration uncertainty and the filter wavelength uncertainty. We include these in our approach, in addition to including HST Calspec calibration uncertainty, 10 SALT2 model systematic uncertainties, a dust systematic, a global redshift bias systematic and also the systematic peculiar velocity uncertainty. This gives thirteen global systematics shared by all surveys, plus two systematics per band in each survey. Denoting $\eta \equiv \lbrace m_B, x_1, c \rbrace$, our initial conditional likelihood for our observed summary statistics shown in Equation \eqref{eq:pop} becomes
\begin{align}
P(\hat{\eta}, \frac{\partial \hat{\eta}}{\partial \Z_i} | \eta, \delta \Z_i, C) = \mathcal{N}(\hat{\eta} + \delta \Z_i \frac{\partial \hat{\eta}}{\partial \Z_i}|\eta,\cov). \label{eq:l3}
\end{align}

\subsubsection{Selection Effects}
\label{sec:selection}
Our treatment of selection effects is to incorporate selection efficiency into our model, rather than relying on simulation-driven data corrections. As such, we need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as
\begin{align}
\mathcal{L} &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As the model so far described in previous sections describe components of a basic likelihood $P(D|\theta)$, and we wish to formulate a function $P(S|{\rm data},\theta)$ that describes the chance of an event being successfully observed, we rearrange the likelihood in terms of those functions and find
\begin{align}
\mathcal{L} &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}
where the denominator represents an integral over all potential data. Full derivation of this can be found in Appendix \ref{app:selection}. As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by HMC (a brief dismissal of months of struggle). We therefore simplify the integral and approximate the selection effects in apparent magnitude and redshift space independently, such that the denominator, denoted now $w$ for simplicity, is given as
\begin{align}
w &= \int  \left[ \int P(S|m_B) P(m_B | z, \theta)\, d m_B \right] P(S|z) P(z|\theta)\, dz. \label{eq:w1}
\end{align}
We apply two further approximations similar to those made in {\rubin} -- that the redshift distribution of the observed supernova reasonably well sampled the $P(S|z)P(z|\theta)$ distribution, and that the survey colour and stretch populations can be treated as Gaussian for the purposes of this integral. The latter assumption is investigated in Appendix \ref{app:approx} and verified to be true for cosmological purposes, if not for robust population determination. The population $P(m_B | z, \theta)$ thus becomes $\mathcal{N}(m_B|m_B^*(z), \sigma^*_{m_B})$, where
\begin{align}
m_B^*(z) &= \langle M_B \rangle + \mu(z) - \alpha \langle x_1(z) \rangle + \beta \langle c(z) \rangle \\
\sigma^*_{m_B} &= \sigma_{M_B}^2 + (\alpha \sigma_{x_1})^2 + (\beta \sigma_c)^2 \notag\\ 
&\quad\quad + 2(\beta \sigma_{M_B,c} - \alpha \sigma{m_B,x_1} - \alpha \beta \sigma_{x_1,c})
\end{align}
What then remains is determining the functional form of $P(S|m_B)$. For the treatment of most surveys, we find that the error function which smoothly transitions from some constant efficiency down to zero is sufficient. Formally, this gives
\begin{align}
P(S|m_B) = \Phi^C(m_B | \mu_{\rm CDF}, \sigma_{\rm CDF}),
\end{align}
where $\Phi^c$ the complimentary CDF and $\mu_{\rm CDF}$ and $\sigma_{\rm CDF}$ specify the selection function. The appropriateness of an error function has been found by many past surveys \citep{Dilday2008, Barbary2010, Perrett2012, Graur2013, Rodney2014}. However, for surveys which suffer from saturation and thus rejection of low-z supernovae, or for groups of surveys treated together (as is common to do with low-redshift surveys), we find that a skew normal is a good analytic form, taking the form
\begin{align}
P(S|m_B) = \mathcal{N}^{{\rm Skew}}(m_B | \mu_{\rm Skew}, \sigma_{\rm Skew}, \alpha_{\rm Skew}).
\end{align}

The selection functions are fit to apparent magnitude efficiency ratios calculated from SNANA simulations, by calculating an efficiency ratio as a function of apparent magnitude. Uncertainty of the Malmquist bias (entering both through statistical uncertainty from finite sized simulations in the efficiency ratio and the discrepancy between the analytic approximation and non-analytic simulation results) is incorporated into the fitting for the analytic approximation. Uncertainty is uniformly added to the efficiency ratio until the reduced $\chi^2$ of the analytic fit reached $1$, allowing us to extract an uncertainty covariance matrix for our analytic fits to either the error function or the skew normal.

With the well sampled approximation as specified previously, we can remove the redshift integral in Eq \eqref{eq:w1} and replace it with a correction for each observed supernova. For the error function (denoted with the subscript `CDF') and skew normal selection functions respectively (denoted with a subscript `Skew'), this correction becomes
\begin{align}
w_{\rm CDF} &= \Phi^c\left(  \frac{m^*_B - \mu_{\rm CDF}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm CDF}^2}}  \right) \label{eq:seldes}\\
w_{\rm Skew} &= 2\mathcal{N}\left( \frac{m_B^* - \mu_{\rm Skew}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm Skew}^2}}\right) \times \notag\\
&\quad\quad \Phi\left(  \frac{{\rm sign}(\alpha_{\rm Skew})(m_B^* - \mu_{\rm Skew})}{\frac{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}{\sigma^2_{\rm Skew}} \sqrt{\frac{\sigma^2_{\rm Skew}}{\alpha^2_{\rm Skew}} + \frac{\sigma_{m_B}^{*2} \sigma^2_{\rm Skew}}{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}} }\right). \label{eq:sellowz}
\end{align}

@@

In order to verify our model we run it through several tests. First, we validate on toy models, verifying that there is not significant cosmological bias in highly constraining datasets. We then validate our model on SNANA simulations based on a collection of low redshift surveys and the DES three year spectroscopic sample.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{simpleApproximateModelOl.pdf}
	\end{center}
	\caption{Posterior surfaces for 200 realisations of supernova data with the $\Lambda$CDM model. Even a large supernova sample when treated robustly is insufficient to provide tight constraints on either $\Omega_m$ and $\Omega_\Lambda$ due to the severe degeneracy between the parameters.}
	\label{fig:simple_ol}
\end{figure}
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{simpleApproximateModelW.pdf}
	\end{center}
	\caption{Posterior surfaces for 200 realisations of supernova data with the Flat $w$CDM model. The well known banana shaped contour is recovered, with the marginalised distributions in $\Omega_m$ and $w$ providing incorrect statistics due to the highly non-Gaussian nature of the posterior surface.}
	\label{fig:simple_w}
\end{figure}

We generate simple toy data to validate the basic premise of the model. For both DES and LowZ data we draw from an underlying $M_B$, $x_1$, $c$ population and translate into apparent magnitude space using $\Omega_m = 0.3$, $\alpha= 0.14$ and $\beta = 3.1$. Masses are randomly drawn from the interval 0 to 1, and a mass correction with $\delta(0) = 0.08$ and $\delta(0)/\delta(\infty) =0.5$ included. Absolute magnitudes are drawn from $\mathcal{N}(-19.3, 0.1)$, true stretch values are drawn from $\mathcal{N}(0, 1)$ and true colour values are drawn from $\mathcal{N}^{\rm Skew}(0, 0.1, 2)$.

Independent observational errors of $0.04$, $0.2$, $0.03$ on $m_B$, $x_1$ and $c$ (following the mean uncertainty for DES SNANA simulations) are added to create the observables. The results are then passed through the selection effects, where each supernova is only selected based on $P(S|m_B)$, using a skew normal function for the LowZ supernovae and error function for the DES-like supernovae. We draw from each survey simulation until we have 1000 LowZ supernovae and 1000 DES-like supernovae, representing a statistical sample of greater power than the estimated 250 supernovae for the DES spectroscopic analysis.

We test four models: Flat $\Lambda$CDM, Flat $w$CDM, $\Lambda$CDM and Flat $w$CDM with a prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$, with the latter included to allow sensitive tests on bias for $w$. To achieve statistical precision, we fit 200 realisations of supernovae datasets. Cosmological parameters are recovered without significant bias. Combined posterior surfaces of all 200 realisations for Flat $wCDM$ fits are shown in Figure \ref{fig:simple_w}. By utilising the Stan framework, fits to these simulations of 2000 supernovae take only on order of a single CPU-hour to run.

To enforce investigate biases in the model in fine detail, we look for systematic bias in $\Omega_m$ in the Flat $\Lambda$CDM cosmology test, and bias in $w$ for the Flat $w$CDM test with strong prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$. This allows us to investigate biases without the investigative hindrances of non-Gaussian or truncated surfaces, and the results of the analysis are detailed in Table \ref{tab:simple_model}, and do not reveal evidence of systematic bias in our model.

\begin{table}
	\centering
	\caption{Maximum likelihood cosmological parameters determined from stacking the surfaces of 200 fits to independent realisations of toy supernova data. As described in the main text, each dataset comprised 1000 low-redshift supernovae and 1000 high-redshift supernovae. Model bias would appear as shifts away from the simulation values of $\Omega_m = 0.3$, $w = -1$. No significant bias is detected in either cosmological model.}
	\label{tab:simple_model}
	\begin{tabular}{ccc}
		\hline
		Model & $\Omega_m$ & $w$ \\ 
		\hline
		Flat $\Lambda$CDM & $0.301\pm 0.016$ & -- \\ 
		Flat $w$CDM + $\Omega_m$ prior & $\left( 300.5^{+10.5}_{-9.8} \right) \times 10^{-3}$ & $-0.998^{+0.043}_{-0.047}$ \\ 
		\hline
	\end{tabular}
\end{table}

@@

Early analyses often treated intrinsic dispersion simply as scatter in the underlying absolute magnitude of the underlying population, but recent analyses require more a more sophisticated approach. In our development of this model and tests of intrinsic dispersion, we analyse the effects of two different scatter models. The first model is the \citet[][hereafter denoted the {\gten} scatter model]{Guy2010}, which models intrinsic scatter with a 70\% contribution from coherent variation and 30\% from chromatic variation. The second model, denoted the {\celeven} model is sourced from \citet{Chotard2011} and has variation with 25\% contribution from coherent scatter and 75\% from chromatic variation.

@@

\begin{table}
	\centering
	\caption{Tested population distributions, where the {\sk} LowZ stretch distribution is formed as sum of two bifurcated Gaussians, with the mean and spread of each component given respectively.}
	\label{tab:dist}
	\begin{tabular}{l|llll}
		\hline
		Model & $\langle x_1 \rangle$ & $\sigma_{x_1}$ &  $\langle c \rangle$ & $\sigma_c$  \\
		\hline
		Gaussian LowZ &  $0.0$ & $1.0$ & $0.0$ & $0.1$ \\
		Gaussian DES  &  $0.0$ & $1.0$ & $0.0$ & $0.1$  \\
		{\sk} LowZ    &  $0.55$ \& $-1.5$ & $^{+0.45}_{-1.0}$ \& $^{+0.5}_{-0.5}$ & $-0.055$ & $^{+0.15}_{-0.023}$ \\
		{\sk} DES     & $0.973$ & $^{+0.222}_{1.472}$ & $-0.054$  &  $^{+0.101}_{0.043}$ \\
		\hline
	\end{tabular}
\end{table}

@@

To forecast the results for the 3 year DES spectroscopic supernova sample, we generate 100 realisations of datasets, each comprising 300 low redshift supernovae and 250 DES-like supernovae. We fit these through the three cosmological models - Flat $\Lambda$CDM, $\Lambda$CDM and $w$CDM, however focus mostly on the $w$CDM model as it is the primary scientific concern of the Dark Energy Survey. We run all three cosmological models with systematics enabled and disabled to explicitly show the uncertainty included from systematic contribution.
